A Monte Carlo Synthetic-Acceleration Method for Solving the Thermal Radiation Diffusion Equation\tnoteref{ornl-cr}

### 2013-10-18 doc/jcomp2012/jcomp2012.tex
Final, final edits.
Final response to editors.

The idea of using Monte Carlo methods (random walks) to invert linear systems
is not new.  The earliest referenced work dates to a 1950 paper by Forsythe
and Leibler \cite{forsythe}.  Their paper credits the idea to unpublished work
by J.v. Neumann and S.M. Ulam dating back to the 1940's.  The basic principles
of Monte Carlo matrix inversion were further elucidated in Hammersley and
Handscomb's 1964 text \cite{hammersley_1964}.  These early methods are
distinguished by very slow, statistically noisy convergence properties; thus,
they have not made any significant impact in the physics community.

@@

To establish the mathematical framework for Monte Carlo linear
solvers, we consider the following matrix equation:
\begin{equation}
  \vA\vx = \vb\:,
  \label{eq:Ax=b}
\end{equation}
which can be written as
\begin{equation}
  \begin{split}
    \vx &= (\vI - \vA)\vx + \vb\\ &= \vH\vx + \vb\:.
  \end{split}
  \label{eq:point_iteration}
\end{equation}
When the spectral radius ($\rho$) of the iteration matrix ($\vH$) is
less than 1, we can expand $\vA^{-1}$ using the \latin{Neumann
  Series}:
\begin{equation}
  \vA^{-1} = (\vI - \vH)^{-1} = \sum_{k=0}^{\infty} \vH^k\:.
\end{equation}
Thus, when $\rho(\vH) < 1$, we can recast the solution vector $\vx$ as
a series:
\begin{equation}
  \begin{split}
    \vx &= (\vI - \vH)^{-1}\vb\\ &= \vb + \vH\vb + \vH^2\vb + \vH^3\vb
    + \ldots\:.
    \label{eq:neumann_series}
  \end{split}
\end{equation}
Using this series, we can write an iterative method that solves
Eq.~(\ref{eq:Ax=b}) (Richardson's Iteration):
\begin{equation}
  \vx^{k+1} = \vH\vx^k + \vb\:.
  \label{eq:richardson}
\end{equation}
Equation~(\ref{eq:richardson}) will converge for all $\vb,\vx^0 \in
\mathcal{R}^{N}$ when $\rho(\vH) < 1$~\cite{kelley_1995}.  All of the Monte
Carlo methods that are described in \S\S~\ref{sec:direct-method} through
\ref{sec:iter-refin-monte} rely on estimating the terms in
Eq.~(\ref{eq:neumann_series}) through random walks.

@@

Now we consider a Monte Carlo method that can be used to estimate the
solution to Eq.~(\ref{eq:Ax=b})~\cite{hammersley_1964}.  First,
rewrite Eq.~(\ref{eq:neumann_series}) in order to calculate a
component of $\vx$:
\begin{equation}
  \begin{split}
    x_i &= (\vb)_i + (\vH\vb)_i + (\vH^2\vb)_i + (\vH^3\vb)_i +
    \ldots\\ &= \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:.
  \end{split}
  \label{eq:neumann_series_i}
\end{equation}
Here, $m$ is the random walk step index and the $h$ terms are the matrix
elements of $\vH$.  The terms in the Neumann series can be interpreted as a
series of transitions from $i_{m-1}\rightarrow i_m$ that can be simulated by a
random walk.  Let $X$ be a random variable sampled from a random walk with $k$
events that initiates in state $i$:
\begin{equation}
  \begin{split}
    X(i_0 = i) &= \sum_{m=0}^{k}W_m b_{i_m}\\ &=
    \sum_{m=0}^{k}w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{m-1},i_m}b_{i_m}\:.
  \end{split}
  \label{eq:estimator_xi}
\end{equation}
Here, the weight on the $m^\text{th}$ step is denoted $W_m$, and each
random walk starts with unit weight. At every step, contributions to
each component of $X$ are generated only in the starting state of the
history, $i_0$, by the source term in the current state,
$b_{i_m}$. Therefore, for a given random walk permutation, each state
must be used as a starting value in order to calculate its component
of the expectation value (i.e. a state of size $N$ will have $N$
histories per random walk permutation). When there is a transition
between states, for example $i\rightarrow j$, the weight is multiplied
by the transition factor
\begin{equation}
  w_{ij} = \frac{h_{ij}}{p_{ij}}\:,
  \label{eq:weight}
\end{equation}
where $p_{ij}$ denotes the probability of transitioning from state $i$
to $j$. Then, the expected value of $X$ is
\begin{equation}
  \begin{split}
    E[X(i_0=i)] &= \sum_{\nu}P_\nu X_\nu\\ &=
    \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N}p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
    w_{i,i_1}w_{i_1,i_2}\dots w_{i_{k-1},i_k}b_{i_k}\\ &= x_i\:,
  \end{split}
  \label{eq:expectation_xi}
\end{equation}
where $\nu$ denotes a particular random walk permutation.  Therefore,
the estimator in Eq.~(\ref{eq:estimator_xi}) is an unbiased estimator
of the components of $\vx$ provided $\rho(\vH) < 1$.

@@

An alternative approach to the one just described is to calculate
contributions to every component of $\vx$ during the random walk. In
this method, the weight change from state $i\rightarrow j$ is
\begin{equation}
  w_{ij} = \frac{h_{ji}}{p_{ij}}\:.
  \label{eq:adjoint-weight}
\end{equation}
The transition probabilities may be calculated as
\begin{equation}
  p_{ij} = \frac{|h_{ji}|}{\sum_{j}|h_{ji}|}\:.
  \label{eq:adjoint-probability}
\end{equation}
Note that the indices are reversed so that the probabilities are normalized
over a column, as opposed to the forward method in which the probabilities are
normalized over a row.  This is equivalent to forming the Neumann series in
reverse order.  Correspondingly, the estimator for this method is
\begin{equation}
  \begin{split}
    X &= \sum_{m=0}^{k}W_m\delta_{i_m,i}\\ &=
    \sum_{m=0}^{k}\hat{b}_{i_0}w_{i_0,i_1}w_{i_1,i_2}\ldots
    w_{i_{m-1},i_m}\delta_{i_m,i}\:.
  \end{split}
  \label{eq:adjoint-tally}
\end{equation}
Here, $\hat{b}_{i_0}$ is the sampled source and initial weight in state $i_0$.
The Kronecker delta implies that tallies are only made in the state where the
random walk currently resides.  Thus, the adjoint method is equivalent to the
common approach in standard Monte Carlo transport simulations. As opposed to
the direct method, the adjoint method does not require starting a history in
each state for each random walk permutation because tallies are made in the
state in which the random walk currently resides. Instead, sampling the source
is sufficient, and one random walk per permutation is required.

@@

A closer examination of the sequential method in Eq.~(\ref{eq:sequential_mc})
shows that this scheme is equivalent to a single preconditioned Richardson
iteration,
\begin{equation}
  \vx^{l+1} = \vx^{l} + \hat{\vA}^{-1}\vr^l\:,
\end{equation}
in which $\vA^{-1}$ indicates the approximate inversion of $\vA$ via the
adjoint Monte Carlo method.  MCSA, on the other hand, is identical to a
Richardson iteration followed by a preconditioned Richardson iteration,
\begin{align}
  \vx^{l+1/2} &= \vx^{l} + \vr^{l}\:,\\
  \vx^{l+1} &= \vx^{l+1/2} + \hat{\vA}^{-1}\vr^{l+1/2}\:.
\end{align}
As above, the adjoint Monte Carlo process is used to approximately invert
$\vA$.

@@

A timing and convergence study is used to demonstrate the effectiveness of the
adjoint method as compared to the direct method. To assess both the CPU time
and number of iterations required to converge to a solution, a problem of
constant $\Delta$ was used with varying number of grid points, fixing the
spectral radius of the system at a constant value for each variation. Both the
five-point and nine-point stencils were used with both the direct and adjoint
solvers. For each case, $N \times N$ total random walk permutations were
computed per MCSA iteration where $N \times N$ is the number of discrete grid
points in the system. Solver parameters were set to a weight cutoff of
\sn{1}{-4} for the stochastic linear solver and a convergence tolerance of
\sn{1}{-8} for the MCSA iterative solver.  Figure~\ref{fig:poisson_cpu_time}
gives the CPU time needed for each case to converge in seconds and
Figure~\ref{fig:poisson_iterations} gives the number of iterations needed for
each case to converge to the specified tolerance as a function of the problem
size. All computations presented in this section and
\S~\ref{subsec:sequential_comparison} were completed on a 3.0 GHz Intel Core 2
Quad Q9650 CPU machine with 16 GB 1067 MHz DDR3 memory.
\begin{figure}[ht!]
  \centering
  \includegraphics[width=5in,clip]{dir_adj_cpu.pdf}
  \caption{\sl CPU Time (s) to converge vs. Problem Size ($N$ for an
    $N \times N$ square mesh). Both the adjoint and direct solvers are
    used with the five point and nine point stencils. A CPU time
    speedup is noted with the adjoint method due to the higher density
    of random walk events in regions with a large residual.}
  \label{fig:poisson_cpu_time}
\end{figure}

@@

We see clearly in Figure~\ref{fig:poisson_cpu_time} that the using the
adjoint solver with MCSA results in a speedup over the direct solver
while the number of iterations required to converge is also reduced as
shown in Figure~\ref{fig:poisson_iterations}. We expect this for several
reasons. First, with an equivalent number of histories specified for
both solvers per MCSA iteration and a system of size $N \times N$, the
direct solver will compute a single random walk for each state in the
system per iteration to acquire a solution in that state, regardless
of the size of the residual in that state. This is necessary in the
direct method to ensure a contribution from each state as the random
walk sequence will only contribute to the starting state. For the
adjoint method, a total of $N \times N$ random walk events will have
their starting state determined by sampling the residual
vector. Because the random walk sequence contributes to the state in
which it currently resides, sampling the residual vector as the Monte
Carlo source gives a higher density of random walk events in regions
with a high residual, thus giving a more accurate correction in that
region due to reduced statistical error. From an iteration
perspective, Figure~\ref{fig:poisson_iterations} shows that using the
direct method yields a roughly unchanging number of iterations
required to converge as the problem size increases. Again, if we
desire a correction value for all states in the problem, then we must
start a random walk in each state in the system which does not reduce
the number of iterations need as the problem size grows. Conversely,
as the problem size grows in the adjoint method, the additional
stochastic histories that will be computed are concentrated in regions
with a large residual, further reducing the stochastic error in the
correction in those regions and subsequently reducing the required
number of iterations to converge.

@@

\subsection{Comparison to Sequential Monte Carlo}
\label{subsec:sequential_comparison}
To further motivate using Monte Carlo Synthetic Acceleration, we compare its
performance to Halton's sequential Monte Carlo method on which our previous
work in this area was based. For this comparison, we use the same transient
Poisson problem as described in the previous section and choose only the
5-point stencil to discretize the Laplacian operator as the previous results
yielded little qualitative difference between the discretizations. Both MCSA
and Halton's method are used with the adjoint Monte Carlo solver. In order to
complete the same study as in the previous section, the number of histories
computed by the Monte Carlo solver at each iteration had to be doubled to $2
\times N \times N$ in order to ensure convergence when using the sequential
Monte Carlo Method. Figure~\ref{fig:seq_cpu_time} gives the CPU time results
for this comparison as a function of problem size while
Figure~\ref{fig:seq_iterations} gives the number of iterations to converge as
a function of problem size. In both cases, using the Monte Carlo solver as a
synthetic acceleration rather than in a sequential Monte Carlo scheme resulted
in a reduction in both CPU time and iterations required to converge. The
additional Richardson iteration between each Monte Carlo solve in the MCSA
method gives a better converged residual source to use with the Monte Carlo
calculation while the sequential method requires more iterations to achieve
the same level of convergence in the residual.

@@

The benefits of using a synthetic acceleration scheme are also noted when the
infinity norm of the residual computed at each iteration for both methods was
collected at each iteration for a fixed problem sizes of $N=100$ and $N=500$
as shown in figures Figure~\ref{fig:seq_100} and \ref{fig:seq_500}
respectively. In both cases, the sequential method is characterized by a slow
approach to the asymptotic rate of convergence.  Conversely, MCSA reaches its
asymptotic convergence rate more quickly and exhibits a faster final rate of
convergence than Halton's method. Even with the doubling of the number of
stochastic histories computed per time step in order to ensure convergence for
the sequential method, we still see robustness issues with a non-monotonically
decreasing residual observed for the $N=100$ case. In both cases the MCSA
solver is observed to be robust with a monotonically decreasing residual.

@@

Using the mathematical preliminaries developed in the previous section, we can
now generate a discrete form of the radiation diffusion equation. We will show
that this discrete form is amenable to interpretation using Monte Carlo
concepts and can be written in a way that a random walk sequence can be
generated.

@@

We choose to use the angle-energy integrated radiation intensity,
$\phi\equiv\phi(\ve{r},t)$
$[\text{GJ}\cdot\text{cm}^{-2}\cdot\text{ns}^{-1}]$, as the primary state
variable.  The intensity is defined as
\begin{equation}
  \label{eq:lte-phi}
  \begin{split}
    \phi(\ve{r},t) &=
    \int_{4\pi}d\bOmega\int_{\nu}d\nu\:\psi(\ve{r},\nu,\hat{\bOmega},t)
    \\ &= \int_{\nu}d\nu\: 4\pi B(\nu,T) \\ &= acT^{4}\:,
  \end{split}
\end{equation}
where $\psi$ is the angular radiation intensity and $\nu$ is the radiation
frequency.  The Planck function (or Planckian) is defined as
\begin{equation}
  B(\nu,T)=\frac{2h\nu^{3}}{c^{2}}\bigl(e^{\frac{h\nu}{kT}} -
  1\bigr)^{-1}\:,
\end{equation}
where $h$ is Planck's constant and $k$ is the Boltzmann constant.  Utilizing
this definition of $\phi$ we can write
\begin{equation}
  \label{eq:partial-T}
  \pder{T}{t} = \frac{1}{4acT^{3}}\pder{\phi}{t}\:.
\end{equation}
Inserting \eqref{eq:partial-T} into \eqref{eq:eq-diff-T} yields
\begin{equation}
  \label{eq:eq-diff}
  \Bigl(\frac{\rho\Cv}{4acT^{3}} + \frac{1}{c}\Bigr)\pder{\phi}{t} -
  \nabla\cdot D\grad{\phi} = Q\:,
\end{equation}
where $D$ is the diffusion coefficient, $D\equiv
D(\ve{r},T(\ve{r},t))\ [\text{cm}]$, and is defined as
\begin{equation}
  D = \frac{1}{3\ros}\:.
\end{equation}
Equation~\eqref{eq:eq-diff} is the model equation we will solve using the
Monte Carlo techniques presented in \S~\ref{sec:monte-carlo-matrix} and
\ref{sec:iter-refin-monte}.

@@

Applying backward-Euler time differencing to Eq.~(\ref{eq:eq-diff}) and
lagging the temperature-dependent coefficients at $t^n$ gives
\begin{equation}
  -\nabla\cdot \Dn\grad{\phi} + \sign\phi = \qn\:,
\end{equation}
where we have suppressed the $n+1$ indices on $\phi$ and
\begin{align}
  \sign &= \frac{\rho \Cv^n}{4ac(\Tn)^3\dt} + \frac{1}{c\dt}\:,\\ \qn
  &= \sign\phin + Q^n\:,\\ \phin &= ac(\Tn)^4\:.
\end{align}
Applying Fick's Law,
\begin{equation}
  \ve{F} = -D\grad{\phi}\:,
  \label{eq:fick}
\end{equation}
where $\ve{F}$ is the radiation flux, we define a flux-balance equation,
\begin{equation}
  \dive{F} + \sign\phi = \qn\:.
  \label{eq:flux-balance}
\end{equation}
Integrating Eq.~(\ref{eq:flux-balance}) over the cell illustrated in
Figure~\ref{fig:cell} gives
\begin{equation}
  \iiint\dive{F}\,d\ve{r} +
  \sign_{ijk}\phi_{ijk}V_{ijk}=\qn_{ijk}V_{ijk}\:,
  \label{eq:flux-balance-vol-integral}
\end{equation}
and $V_{ijk} = \Delta_i\Delta_j\Delta_k$.
\begin{figure}[ht!]
  \centerline{ \includegraphics[clip]{mesh_cell.pdf} }
  \caption{Orthogonal 3D mesh cell.}
  \label{fig:cell}
\end{figure}
We now apply the divergence theorem,
\begin{equation}
  \begin{split}
    \iiint\dive{F}\,d\ve{r} &= \oint\hat{\ve{n}}\cdot\ve{F}\,dA\\ &=
    \sum_{f=1}^{6}\hat{\ve{n}}_f\cdot\ve{F}_f A_f\:,
  \end{split}
\end{equation}
to Eq.(\ref{eq:flux-balance-vol-integral}) yielding
\begin{equation}
  \begin{split}
    (F_{i+1/2} - F_{i-1/2})\Dj\Dk &+ (F_{j+1/2} - F_{j-1/2})\Di\Dk +
    (F_{k+1/2}\\ &- F_{k-1/2})\Di\Dj + \sign_{ijk}\phi_{ijk}V_{ijk} =
    \qn_{ijk}V_{ijk}\:.
  \end{split}
  \label{eq:flux-balance-difference}
\end{equation}

>>>>

The idea of using Monte Carlo methods (random walks) to invert linear systems
is not new.  The earliest referenced work dates to a 1950 paper by Forsythe
and Leibler \cite{forsythe}.  Their paper credits the idea to unpublished work
by J. von Neumann and S.M. Ulam dating back to the 1940's.  The basic
principles of Monte Carlo matrix inversion were further elucidated in
Hammersley and Handscomb's 1964 text \cite{hammersley_1964}.  These early
methods are distinguished by very slow, statistically noisy convergence
properties; thus, they have not made any significant impact in the physics
community.

@@

To establish the mathematical framework for Monte Carlo linear
solvers, we consider the following matrix equation:
\begin{equation}
  \vA\vx = \vb\:,
  \label{eq:Ax=b}
\end{equation}
which can be written as
\begin{equation}
  \begin{split}
    \vx &= (\vI - \vA)\vx + \vb\\ &= \vH\vx + \vb\:.
  \end{split}
  \label{eq:point_iteration}
\end{equation}
When the spectral radius ($\rho$) of the iteration matrix ($\vH$) is
less than 1, we can expand $\vA^{-1}$ using the \latin{Neumann
  Series}:
\begin{equation}
  \vA^{-1} = (\vI - \vH)^{-1} = \sum_{k=0}^{\infty} \vH^k\:.
\end{equation}
Thus, when $\rho(\vH) < 1$, we can recast the solution vector $\vx$ as
a series:
\begin{equation}
  \begin{split}
    \vx &= (\vI - \vH)^{-1}\vb\\ &= \vb + \vH\vb + \vH^2\vb + \vH^3\vb
    + \ldots\:.
    \label{eq:neumann_series}
  \end{split}
\end{equation}
Using this series, we can write an iterative method, known as Richardson's
Iteration, that solves Eq.~(\ref{eq:Ax=b}):
\begin{equation}
  \vx^{k+1} = \vH\vx^k + \vb\:.
  \label{eq:richardson}
\end{equation}
Equation~(\ref{eq:richardson}) will converge for all $\vb,\vx^0 \in
\mathcal{R}^{N}$ when $\rho(\vH) < 1$~\cite{kelley_1995}.  All of the Monte
Carlo methods that are described in \S\S~\ref{sec:direct-method} through
\ref{sec:iter-refin-monte} rely on estimating the terms in
Eq.~(\ref{eq:neumann_series}) through random walks.

@@

Now we consider a Monte Carlo method that can be used to estimate the
solution to Eq.~(\ref{eq:Ax=b})~\cite{hammersley_1964}.  First,
rewrite Eq.~(\ref{eq:neumann_series}) in order to calculate a
component of $\vx$:
\begin{equation}
  \begin{split}
    x_i &= (\vb)_i + (\vH\vb)_i + (\vH^2\vb)_i + (\vH^3\vb)_i +
    \ldots\\ &= \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:.
  \end{split}
  \label{eq:neumann_series_i}
\end{equation}
Here, $k$ is the random walk step index and the $h$ terms are the matrix
elements of $\vH$.  The terms in the Neumann series can be interpreted as a
series of transitions from $i_{k-1}\rightarrow i_k$ that can be simulated by a
random walk.  Let $X$ be a random variable sampled from a random walk with $k$
events that initiates in state $i$:
\begin{equation}
  \begin{split}
    X(i_0 = i) &= \sum_{m=0}^{k}W_m b_{i_m}\\ &=
    \sum_{m=0}^{k}w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{m-1},i_m}b_{i_m}\:.
  \end{split}
  \label{eq:estimator_xi}
\end{equation}
Here, the weight on the $m^\text{th}$ step is denoted $W_m$, and each
random walk starts with unit weight. At every step, contributions to
each component of $X$ are generated only in the starting state of the
history, $i_0$, by the source term in the current state,
$b_{i_m}$. Therefore, for a given random walk permutation, each state
must be used as a starting value in order to calculate its component
of the expectation value (i.e. a state of size $N$ will have $N$
histories per random walk permutation). When there is a transition
between states, for example $i\rightarrow j$, the weight is multiplied
by the transition factor
\begin{equation}
  w_{ij} = \frac{h_{ij}}{p_{ij}}\:,
  \label{eq:weight}
\end{equation}
where $p_{ij}$ denotes the probability of transitioning from state $i$
to $j$. Then, the expected value of $X$ is
\begin{equation}
  \begin{split}
    E[X(i_0=i)] &= \sum_{\nu}P_\nu X_\nu\\ &=
    \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N}p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
    w_{i,i_1}w_{i_1,i_2}\dots w_{i_{k-1},i_k}b_{i_k}\\ &= x_i\:,
  \end{split}
  \label{eq:expectation_xi}
\end{equation}
where $\nu$ denotes a particular random walk permutation.  Therefore,
the estimator in Eq.~(\ref{eq:estimator_xi}) is an unbiased estimator
of the components of $\vx$ provided $\rho(\vH) < 1$.

@@

An alternative approach to the direct method is to calculate contributions to
every component of $\vx$ during the random walk. In this method, the weight
change from state $i\rightarrow j$ is
\begin{equation}
  w_{ij} = \frac{h_{ji}}{p_{ij}}\:.
  \label{eq:adjoint-weight}
\end{equation}
The transition probabilities may be calculated as
\begin{equation}
  p_{ij} = \frac{|h_{ji}|}{\sum_{j}|h_{ji}|}\:.
  \label{eq:adjoint-probability}
\end{equation}
Note that the indices are reversed so that the probabilities are normalized
over a column, as opposed to the forward method in which the probabilities are
normalized over a row.  This is equivalent to forming the Neumann series in
reverse order.  Correspondingly, the estimator for this method is
\begin{equation}
  \begin{split}
    X &= \sum_{m=0}^{k}W_m\delta_{i_m,i}\\ &=
    \sum_{m=0}^{k}\hat{b}_{i_0}w_{i_0,i_1}w_{i_1,i_2}\ldots
    w_{i_{m-1},i_m}\delta_{i_m,i}\:.
  \end{split}
  \label{eq:adjoint-tally}
\end{equation}
Here, $\hat{b}_{i_0}$ is the sampled source and initial weight in state $i_0$.
The Kronecker delta implies that tallies are only made in the state where the
random walk currently resides.  Thus, the adjoint method is equivalent to the
common approach in standard Monte Carlo transport simulations. As opposed to
the direct method, the adjoint method does not require starting a history in
each state for each random walk permutation because tallies are made in the
state in which the random walk currently resides. Instead, sampling the source
is sufficient, and one random walk per permutation is required.

@@

A closer examination of the sequential method in Eq.~(\ref{eq:sequential_mc})
shows that this scheme is equivalent to a single preconditioned Richardson
iteration,
\begin{equation}
  \vx^{l+1} = \vx^{l} + \hat{\vA}^{-1}\vr^l\:,
\end{equation}
in which $\hat{\vA}^{-1}$ indicates the approximate inversion of $\vA$ via the
adjoint Monte Carlo method.  MCSA, on the other hand, is identical to a
Richardson iteration followed by a preconditioned Richardson iteration,
\begin{align}
  \vx^{l+1/2} &= \vx^{l} + \vr^{l}\:,\\
  \vx^{l+1} &= \vx^{l+1/2} + \hat{\vA}^{-1}\vr^{l+1/2}\:.
\end{align}
As above, the adjoint Monte Carlo process is used to approximately invert
$\vA$.

@@

A timing and convergence study is used to demonstrate the improved
effectiveness of the adjoint method as compared to the direct method. To
assess both the CPU time and number of iterations required to converge to a
solution, a problem of constant $\Delta$ was used with varying number of grid
points, fixing the spectral radius of the system at a constant value for each
variation. Both the five-point and nine-point stencils were used with both the
direct and adjoint solvers. For each case, $N \times N$ total random walk
permutations were computed per MCSA iteration where $N \times N$ is the number
of discrete grid points in the system. Solver parameters were set to a weight
cutoff of \sn{1}{-4} for the stochastic linear solver and a convergence
tolerance of \sn{1}{-8} for the MCSA iterative solver.
Figure~\ref{fig:poisson_cpu_time} gives the CPU time needed for each case to
converge in seconds and Figure~\ref{fig:poisson_iterations} gives the number
of iterations needed for each case to converge to the specified tolerance as a
function of the problem size. All computations presented in this section and
\S~\ref{subsec:sequential_comparison} were completed on a 3.0 GHz Intel Core 2
Quad Q9650 CPU machine with 16 GB 1067 MHz DDR3 memory.
\begin{figure}[ht!]
  \centering
  \includegraphics[width=5in,clip]{dir_adj_cpu.pdf}
  \caption{\sl CPU Time (s) to converge vs. Problem Size ($N$ for an
    $N \times N$ square mesh). Both the adjoint and direct solvers are
    used with the five point and nine point stencils. A CPU time
    speedup is achieved with the adjoint method due to the higher density
    of random walk events in regions with a large residual.}
  \label{fig:poisson_cpu_time}
\end{figure}

@@

We see in Figures~\ref{fig:poisson_cpu_time} and \ref{fig:poisson_iterations}
that the adjoint solver with MCSA is faster and requires fewer iterations than
the direct method. Several reasons explain this result. First, with an
equivalent number of histories specified for both solvers per MCSA iteration
and a system of size $N \times N$, the direct solver will compute a single
random walk for each state in the system per iteration to acquire a solution
in that state, regardless of the size of the residual in that state. This is
necessary in the direct method to ensure a contribution from each state as the
random walk sequence will only contribute to the starting state. For the
adjoint method, a total of $N \times N$ random walk events will have their
starting state determined by sampling the residual vector. Because the random
walk sequence contributes to the state in which it currently resides, sampling
the residual vector as the Monte Carlo source gives a higher density of random
walk events in regions with a high residual, thus giving a more effective
correction in those regions due to reduced statistical error. From an
iteration perspective, Figure~\ref{fig:poisson_iterations} shows that using
the direct method yields a roughly unchanging number of iterations required to
converge as the problem size increases. Again, if we desire a correction value
for all states in the problem, then we must start a random walk in each state
in the system which does not reduce the number of iterations needed as the
problem size grows. Conversely, as the problem size grows in the adjoint
method, the additional stochastic histories that will be computed are
concentrated in regions with a large residual, further reducing the stochastic
error in the correction in those regions and ultimately reducing the required
number of iterations to converge.

@@

\subsection{Comparison to Sequential Monte Carlo}
\label{subsec:sequential_comparison}
To further motivate using Monte Carlo Synthetic Acceleration, we compare its
performance to Halton's sequential Monte Carlo method on which our previous
work in this area was based. For this comparison, we use the same transient
Poisson problem as described in the previous section and choose only the
5-point stencil to discretize the Laplacian operator because the previous
results yielded little qualitative difference between the
discretizations. Both MCSA and Halton's method are used with the adjoint Monte
Carlo solver. In order to complete the same study as in the previous section,
the number of histories computed by the Monte Carlo solver at each iteration
had to be doubled to $2 \times N \times N$ in order to ensure convergence when
using the sequential Monte Carlo Method. Figure~\ref{fig:seq_cpu_time} gives
the CPU time results for this comparison as a function of problem size, and
Figure~\ref{fig:seq_iterations} gives the number of iterations to converge as
a function of problem size. In both cases, using the Monte Carlo solver as a
synthetic acceleration rather than in a sequential Monte Carlo scheme resulted
in a reduction in both CPU time and iterations required to converge. The
additional Richardson iteration between each Monte Carlo solve in the MCSA
method gives a better converged residual source to use with the Monte Carlo
calculation while the sequential method requires more iterations to achieve
the same level of convergence in the residual.

@@

The benefits of using a synthetic acceleration scheme are also illustrated
when comparing the residuals at each iteration.  Figures~\ref{fig:seq_100} and
\ref{fig:seq_500} show the infinity norm of the residual computed at each
iteration for both methods for fixed problems sizes of $N=100$ and $500$,
respectively.  In both cases, the sequential method is characterized by a slow
approach to the asymptotic rate of convergence.  Conversely, MCSA reaches its
asymptotic convergence rate more quickly and exhibits a faster final rate of
convergence than Halton's method. Even with the doubling of the number of
stochastic histories computed per time step in order to ensure convergence for
the sequential method, we still see robustness issues with a non-monotonically
decreasing residual observed for the $N=100$ case. In both cases the MCSA
solver is observed to be robust with a monotonically decreasing residual.

@@

Using the mathematical preliminaries developed in the previous section, we can
now generate a discrete form of the radiation diffusion equation. We will show
that this discrete form is amenable to interpretation using Monte Carlo
concepts and can be written in a way that generates a random walk sequence.

@@

We choose to use the angle- and energy-integrated radiation intensity,
$\phi\equiv\phi(\ve{r},t)$
$[\text{GJ}\cdot\text{cm}^{-2}\cdot\text{ns}^{-1}]$, as the primary state
variable.  The intensity is defined as
\begin{equation}
  \label{eq:lte-phi}
  \begin{split}
    \phi(\ve{r},t) &=
    \int_{4\pi}d\bOmega\int_{\nu}d\nu\:\psi(\ve{r},\nu,\hat{\bOmega},t)
    \\ &= \int_{\nu}d\nu\: 4\pi B(\nu,T) \\ &= acT^{4}\:,
  \end{split}
\end{equation}
where $\psi$ is the angular radiation intensity and $\nu$ is the radiation
frequency.  The Planck function (or Planckian) is defined as
\begin{equation}
  B(\nu,T)=\frac{2h\nu^{3}}{c^{2}}\bigl(e^{\frac{h\nu}{kT}} -
  1\bigr)^{-1}\:,
\end{equation}
where $h$ is Planck's constant and $k$ is the Boltzmann constant.  Utilizing
this definition of $\phi$ we can write
\begin{equation}
  \label{eq:partial-T}
  \pder{T}{t} = \frac{1}{4acT^{3}}\pder{\phi}{t}\:.
\end{equation}
Inserting \eqref{eq:partial-T} into \eqref{eq:eq-diff-T} yields
\begin{equation}
  \label{eq:eq-diff}
  \Bigl(\frac{\rho\Cv}{4acT^{3}} + \frac{1}{c}\Bigr)\pder{\phi}{t} -
  \nabla\cdot D\grad{\phi} = Q\:,
\end{equation}
where $D$ is the diffusion coefficient, $D\equiv
D(\ve{r},T(\ve{r},t))\ [\text{cm}]$, and is defined as
\begin{equation}
  D = \frac{1}{3\ros}\:.
\end{equation}
Equation~\eqref{eq:eq-diff} is the model equation we will solve using the
Monte Carlo techniques presented in \S~\ref{sec:monte-carlo-matrix} and
\ref{sec:iter-refin-monte}.

@@

Applying backward-Euler time differencing to Eq.~(\ref{eq:eq-diff}) and
lagging the temperature-dependent coefficients at $t^n$ gives
\begin{equation}
  -\nabla\cdot \Dn\grad{\phi} + \sign\phi = \qn\:,
  \label{eq:backwards_euler}
\end{equation}
where we have suppressed the $n+1$ indices on $\phi$ and
\begin{align}
  \sign &= \frac{\rho \Cv^n}{4ac(\Tn)^3\dt} + \frac{1}{c\dt}\:,\\ \qn
  &= \sign\phin + Q^n\:,\\ \phin &= ac(\Tn)^4\:.
\end{align}
Applying Fick's Law,
\begin{equation}
  \ve{F} = -D\grad{\phi}\:,
  \label{eq:fick}
\end{equation}
where $\ve{F}$ is the radiation flux, we define a flux-balance equation,
\begin{equation}
  \dive{F} + \sign\phi = \qn\:.
  \label{eq:flux-balance}
\end{equation}
Integrating Eq.~(\ref{eq:flux-balance}) over the cell illustrated in
Figure~\ref{fig:cell} gives
\begin{equation}
  \iiint\dive{F}\,d\ve{r} +
  \sign_{ijk}\phi_{ijk}V_{ijk}=\qn_{ijk}V_{ijk}\:,
  \label{eq:flux-balance-vol-integral}
\end{equation}
and $V_{ijk} = \Delta_i\Delta_j\Delta_k$.
\begin{figure}[ht!]
  \centerline{ \includegraphics[clip]{mesh_cell.pdf} }
  \caption{Orthogonal 3D mesh cell.}
  \label{fig:cell}
\end{figure}
We now apply the divergence theorem,
\begin{equation}
  \begin{split}
    \iiint\dive{F}\,d\ve{r} &= \oint\hat{\ve{n}}\cdot\ve{F}\,dA\\ &=
    \sum_{f=1}^{6}\hat{\ve{n}}_f\cdot\ve{F}_f A_f\:,
  \end{split}
\end{equation}
to Eq.(\ref{eq:flux-balance-vol-integral}) yielding
\begin{equation}
  \begin{split}
    (F_{i+1/2} - F_{i-1/2})\Dj\Dk &+ (F_{j+1/2} - F_{j-1/2})\Di\Dk +
    (F_{k+1/2}\\ &- F_{k-1/2})\Di\Dj + \sign_{ijk}\phi_{ijk}V_{ijk} =
    \qn_{ijk}V_{ijk}\:.
  \end{split}
  \label{eq:flux-balance-difference}
\end{equation}
